The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-10-04 paper.tex
specify that German DL Syntax is used for axioms
start submission nr 4

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
\usepackage{tablefootnote}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier
\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\newcommand{\latin}[1]{\emph{#1}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

\paragraph{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
The FMA primarily focuses on depicting the entirety of human anatomy.
This broad approach results in inconsistencies in how skeletal structures are detailed, as the FMA's overarching goal to cover anatomy in general cannot sufficiently address the complexities and nuances of skeletal anatomy.
Such a wide scope also means that the anatomical knowledge pertaining to the skeleton is not depicted with the depth or precision that anthropological work demands, reflecting the limitations in existing literature on the subject.
From an anthropological perspective, the FMA ontology exhibits challenges in terms of content depth, comprehensiveness, and granularity.
Specifically, the main issues encompass a lack of cross-references between concepts, taxonomic and hierarchical inconsistencies, insufficient detail levels, and conceptual ambiguity.
These gaps result in a notable absence of detailed relationships, associations, and concepts that are crucial for a thorough anthropological understanding and analysis.
Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
%
Interlinks from ANNO to the FMA are added manually as it requires decisions from the domain experts.
The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.

@@

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom:
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
%
The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
In both cases, the consistency of the ontology was proven.
Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, 14 of the core categories%
\footnote{Anatomical line, Anatomical point, Anatomical structure, Bone, Bone compound, Bone part, Measurement point, Orientation point, Phenotype, Spatial anatomical entity, Tooth, Tooth part, Phenotype and Relative anatomical location.}
and 4 of the relation types%
\footnote{hasPart, boundaryOf, locationOf and derivedFrom.}
of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
Statistical information about the ontology is shown in \cref{tab:stats}.

@@

The integration of ANNO into \aw{} significantly enhances its functionality.
This integration involves four steps, the first of which was importing the ontology data in the form of JSON files.
The application structure is then adjusted to align with the attributes defined in the ontology.
This process includes assigning objects within the application, such as markers and measurements, to corresponding objects in the ontology.
Furthermore, the ontology import process in \aw{} from a JSON file involves organizing the information hierarchically from this file and adding application-specific details, such as spatial positioning data in 3D space and placeholder models.
These additional details are stored in a separate JSON file, ensuring compatibility with newer ontology versions.
During runtime, \aw{} interprets the imported information, creating containers for the bone data to be imported and corresponding entries in the forms.
The second step involved the adaptation of the properties of the bone objects and the related user interface elements, as the ontology includes attributes for these objects not previously implemented in \aw{}.
This has led to modifications in the input forms and lists to accommodate these new attributes.
As a third step, the application was adapted to facilitate the assignment of measurements and markings to concepts from the ontology.
For instance, measurements taken within the application can be mapped to predefined measurement paths in the ontology.
To achieve this the users can choose from a list of features from the ontology to be assigned to their measurements and markings.
Lastly, the import of discriminant functions allowed for their interpretation within the application.
This includes a form for selecting measurement paths relevant to a particular discriminant function and the computation of these functions based on measurements created and assigned by users.
This feature notably enhances the application's capability for tasks such as sex determination.

>>>>

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
\usepackage{tablefootnote}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier
\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\newcommand{\latin}[1]{\emph{#1}}
% uncomment exactly one of the following two lines:
\newcommand{\revision}[1]{\textcolor{red}{#1}}% for seeing changes in red
%\newcommand{\revision}[1]{#1}% for final output
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

\paragraph{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
The FMA primarily focuses on depicting the entirety of human anatomy.
This broad approach results in inconsistencies in how skeletal structures are detailed, as the FMA's overarching goal to cover anatomy in general cannot sufficiently address the complexities and nuances of skeletal anatomy.
Such a wide scope also means that the anatomical knowledge pertaining to the skeleton is not depicted with the depth or precision that anthropological work demands, reflecting the limitations in existing literature on the subject.
From an anthropological perspective, the FMA ontology exhibits challenges in terms of content depth, comprehensiveness, and granularity.
Specifically, the main issues encompass a lack of cross-references between concepts, taxonomic and hierarchical inconsistencies, insufficient detail levels, and conceptual ambiguity.
These gaps result in a notable absence of detailed relationships, associations, and concepts that are crucial for a thorough anthropological understanding and analysis.
Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
%
\revision{%
Interlinks from ANNO to the FMA are added manually as decisions by the domain experts are required because seemingly equivalent classes may be hypernyms, hyponyms or not related at all.
}%
The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.

@@

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom (given in German DL Syntax):
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
%
The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
In both cases, the consistency of the ontology was proven.
\revision{%
Independent of the proof of consistency by reasoners, no inconsistencies were to be expected for two reasons:
Firstly, the overall ontology is based on a core component (ANNOdc), which has a clear easy-to-understand structure and was developed by ontology experts.
Secondly, the methodology we have established for developing domain-specific ontologies based on core ontologies does not allow any consistency violations, see \cref{sec:annods}.
}

Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, 14 of the core categories%
\footnote{Anatomical line, Anatomical point, Anatomical structure, Bone, Bone compound, Bone part, Measurement point, Orientation point, Phenotype, Spatial anatomical entity, Tooth, Tooth part, Phenotype and Relative anatomical location.}
and 4 of the relation types%
\footnote{hasPart, boundaryOf, locationOf and derivedFrom.}
of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
Statistical information about the ontology is shown in \cref{tab:stats}.

@@

The integration of ANNO into \aw{} significantly enhances its functionality.
This integration involves four steps, the first of which was importing the ontology data in the form of JSON files.
The application structure is then adjusted to align with the attributes defined in the ontology.
This process includes assigning objects within the application, such as markers and measurements, to corresponding objects in the ontology.
Furthermore, the ontology import process in \aw{} from a JSON file involves organizing the information hierarchically from this file and adding application-specific details, such as spatial positioning data in 3D space and placeholder models.
These additional details are stored in a separate JSON file, ensuring compatibility with newer ontology versions.
During runtime, \aw{} interprets the imported information, creating containers for the bone data to be imported and corresponding entries in the forms.
The second step involved the adaptation of the properties of the bone objects and the related user interface elements, as the ontology includes attributes for these objects not previously implemented in \aw{}.
This has led to modifications in the input forms and lists to accommodate these new attributes.
As a third step, the application was adapted to facilitate the assignment of measurements and markings to concepts from the ontology.
For instance, measurements taken within the application can be mapped to predefined measurement paths in the ontology.
To achieve this the users can choose from a list of features from the ontology to be assigned to their measurements and markings.

Lastly, the import of discriminant functions allowed for their interpretation within the application.
\revision{%
This updates the previous workflow of manually recording the results of the measurements, calculating the discriminant function outside of our application and evaluating the result manually to just selecting the relevant discriminant function, checking if the measurements are correctly assigned to the parts of the function and confirming their input whereupon the resulting classification is shown on the screen.
}
This feature notably enhances the application's capability for tasks such as sex determination.
